---
title: "PCR_positivity"
author: "ncov-RTM Imperial College London"
date: "27/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE}
packages <- c("VGAM","ggplot2","rriskDistributions",
              "survival","survminer","gsci","flexsurv","stats")
lapply(packages, require, character.only = TRUE)
df <- read.csv("omar_data.csv")
df <- df[,1:4]
df <- subset(df,!is.na(Days))
```

# Time from symptom onset to PCR negative

In *[Eurosurveillance 2020 Jul 30; 25(30): 2001292](https://dx.doi.org/10.2807/1560-7917.ES.2020.25.30.2001292)*, Omar et al. analysed data of consecutive RT-qPCR results in 537 symptomatic patients with mild COVID-19 in home quarantine, between 28 February to 6 June 2020 in Rhineland-Palatinate, Germany. Table 1 presents their reported data on the population with an RT-qPCR positive for SARS-CoV-2 RNA at mean (95%CI) days after symptom onset. They reported a **median time to testing negative of 20 days (IQR 16-28)**.

### Table 1. Cumulative population proportions, from *[Omar et al.](https://dx.doi.org/10.2807/1560-7917.ES.2020.25.30.2001292)*

```{r, echo=FALSE}
df
```

Using Omar et al.'s reported cumulative `proportion`  and `mean` time, we fitted a gamma accelerated failure time regression line. 

```{r,include = FALSE}
# fit parameters directly
p <- rev(df$Prop[!is.na(df$Proportion)])
q <- df$Days[!is.na(df$Proportion)]

fit.results <- rriskFitdist.perc(p, q, show.output = FALSE)
params <- as.numeric(get.gamma.par(p,q),show.output = FALSE, plot = FALSE)
edata <- data.frame(y = rgamma(1000, shape=params[1], scale=1/params[2]))

# Fit gamma-accelerated failure time model
edata.surv <- Surv(edata$y)
fit.edata <- survfit(edata.surv~1,type="kaplan-meier", conf.type="log-log")

```
```{r, echo=FALSE}
print(fit.edata,print.rmean=T)
ggsurvplot(fit.edata,data=edata,
           palette="lancet",
           ggtheme=theme_bw(),
           surv.median.line="hv",
           legend.title="Time to SARS-CoV-2 negative",
           legend.labs="",xlab="Days since symptoms onset",
           ylab="Probability still positive")
fit.gamma.edata <- flexsurvreg(edata.surv~1, dist="gamma")
fit.gamma.edata
```

## Fit Erlang distribution 

From the gamma accelerated failure function, we extract the cumulative distribution function over a sequence of survival probability of length 21 (*e.g. 0, 0.05, 0.10, ... 0.95, 0.99*). That is, the likelihood of remaining **PCR positive over scaled observation time** (*dd*) and the mean, median and IQR of progression duration time (unscaled). 

We used these data to fit an *Erlang* (`k,y`) distribution of time from symptom onset to PCR negativity, by minimizing the Kullback-Leibler divergence. 

```{r, echo = FALSE}
# Obtain ccdf of PCR positivity
q <- seq(0, 1000, l = 21)
df.edata <- data.frame(x = fit.edata$surv, y = fit.edata$cumhaz, 
                       z = fit.edata$time)
ccdf_pcr <- data.frame()
for (i in 1:21) {
  if (i == 1){out = 0}
  else if (i>1){
    x = q[i]
    out = df.edata$y[x]
  }
  ccdf_pcr[1,i] <- out
}
rownames(ccdf_pcr) <- "pcr_pos"

# obtain mean, median and IQR of progression durations
progression_durations <- data.frame(param = "m_PosNeg",
                                    Q2.5 = df.edata$z[25],
                                    Q50 = df.edata$z[500],
                                    Q97.5 = df.edata$z[974],
                                    mean = mean(df.edata$z))

# pass parameters on to Lilith's functions
source("fit_Erlang.R") ## this is kept offline
```
```{r,}
op_res # scaled Erlang parameters (time in dd)
scaled_gamma # re-scaled gamma over mean duration (true time)
progression_durations$mean # mean duration
```
