---
title: "odin"
author: "Pablo Perez"
date: "30/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rich's odin vignette

`odin` implements a high-level language for describing and implementing ordinary differential equations in R. It provides a "domain specific language" (DSL) which looks like R but is compiled directly to C. The actual solution of the differential equations is done with the `deSolve` package.

This example is the Lorenz attractor implemented in `odin`, which's main purpose is to simplify the transition from R to C language. This way, ODE models can be written and solved quickly.  


```{r}
lorenz <- odin::odin({
  ## Derivatives
  deriv(y1) <- sigma * (y2 - y1)
  deriv(y2) <- R * y1 - y2 - y1 * y3
  deriv(y3) <- -b * y3 + y1 * y2

  ## Initial conditions
  initial(y1) <- 10.0
  initial(y2) <- 1.0
  initial(y3) <- 1.0

  ## parameters
  sigma <- 10.0
  R     <- 28.0
  b     <-  8.0 / 3.0
})

mod <- lorenz()
t <- seq(0, 100, length.out = 50000)
y <- mod$run(t)
```

## Including Plots


