---
title: "RT-PCR positivity duration"
author: "Pablo Perez"
date: "19/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("VGAM","ggplot2","rriskDistributions",
              "survival","survminer","gsci","flexsurv","stats")
lapply(packages, require, character.only = TRUE)
df <- read.csv("omar_data.csv")
```

In *[Eurosurveillance 2020 Jul 30; 25(30): 2001292](https://dx.doi.org/10.2807/1560-7917.ES.2020.25.30.2001292)*, Omar et al. analysed data of consecutive RT-qPCR results in 537 symptomatic patients with mild COVID-19 in home quarantine, between 28 February to 6 June 2020 in Rhineland-Palatinate, Germany. Table 1 presents their reported data on the population with an RT-qPCR positive for SARS-CoV-2 RNA at mean (95%CI) days after symptom onset. They reported a **median time to testing negative of 20 days (IQR 16-28)**.

### Table 1. Cumulative population proportions

```{r, echo=TRUE}
df[,2:4]
```

## Simulated study population
For `i in 1:1,000`, at each iteration of time `t`, sample `n` random numbers $$\ n \ = \space n_{t} \ - \ n_{t-1}$$ from a normal distribution with mean = `mu` and sd = `sigma`, given upper and lower bounds of observed time.

```{r, include = FALSE}
N <- 1000
dist.data <- matrix(nrow=N,ncol=2)

for (i in 1:N) {
  while (TRUE) {
    edata <- NULL
    for (t in 1:12) {
      
      n <- df$n[t]
      out <- data.frame(x = rnorm(n, mean=df$mu[t], sd=df$sigma[t]))
      edata <- rbind(edata,out)
    }
    
    fit <- try(fitdistrplus::fitdist(edata$x,"gamma",lower=c(0,0),method = "mle",
                                     start = list(scale=0.1,shape=0.1)),silent=T)
    if(!is(fit, 'try-error')) break
  }
  dist.data[i,1] <- fit$estimate[1]
  dist.data[i,2] <- fit$estimate[2]
  rm(edata)
}
edata <- data.frame(y = rgamma(100, shape=mean(dist.data[,2]), rate=1/mean(dist.data[,1])))
```

At each step `i`, fit a gamma distribution by maximum likelihood, obtain its parameters and then average across all 1,000 iterations. 

```{r, echo=FALSE}
print(c("Shape = ",mean(dist.data[,2])))
print(c("Scale = ",mean(dist.data[,1])))
print(c("Rate = ",1/mean(dist.data[,1])))
fitdistrplus::plotdist(edata$y, histo=T, demp=T)
summary(edata$y)
```

## Fit distribution directly 

Using Omar et al.'s reported cumulative `proportion`  and `mean` time, fit distribution lines and obtain parameters. 

```{r,include = FALSE}
p <- rev(df$Prop[!is.na(df$Proportion)])
q <- df$Days[!is.na(df$Proportion)]

fit.results <- rriskFitdist.perc(p, q, show.output = FALSE)
params <- as.numeric(get.gamma.par(p,q),show.output = FALSE, plot = FALSE)
```

```{r,echo=TRUE}
get.gamma.par(p, q, show.output = FALSE, plot = TRUE, tol = 0.001)

```
Obtain a gamma distributed random sample of `n = 1000` and test distribution fit. 

```{r,include = FALSE}
edata.2 <- data.frame(y = rgamma(1000, shape=params[1], scale=1/params[2]))
```
```{r,echo=FALSE}
fitdistrplus::plotdist(edata.2$y, histo=T, demp=T)
summary(edata.2$y)
```


## Fit gamma accelerated failure time regression
Using simulated datasets above, let's fit a gamma accelerated failure time regression model and obtain parameters. 
```{r, include=FALSE}
edata.surv <- Surv(edata$y)
edata.2.surv <- Surv(edata.2$y)

fit.edata <- survfit(edata.surv~1,type="kaplan-meier", conf.type="log-log")
fit.edata.2 <- survfit(edata.2.surv~1,type="kaplan-meier", conf.type="log-log")
```
### Figure 1. Simulated population
```{r}
print(fit.edata,print.rmean=T)
ggsurvplot(fit.edata,data=edata,
           palette="lancet",
           ggtheme=theme_bw(),
           surv.median.line="hv",
           legend.title="Time to SARS-CoV-2 negative",
           legend.labs="",xlab="Days since symptoms onset",
           ylab="Probability still positive")
fit.gamma.edata <- flexsurvreg(edata.surv~1, dist="gamma")
fit.gamma.edata
```
### Figure 2. Direct fit
```{r}
print(fit.edata.2,print.rmean=T)
ggsurvplot(fit.edata.2,data=edata.2,
           palette="lancet",
           ggtheme=theme_bw(),
           surv.median.line="hv",
           legend.title="Time to SARS-CoV-2 negative",
           legend.labs="",xlab="Days since symptoms onset",
           ylab="Probability still positive")
fit.gamma.edata.2 <- flexsurvreg(edata.2.surv~1, dist="gamma")
fit.gamma.edata.2
```

## Fit Erlang distribution 

From the gamma accelerated failure function fitted to simulated data `edata`, extract the cumulative distribution function `ccdf_pcr` parameters over a sequence of survival probability of length 21 (*e.g. 0, 0.05, 0.10, ... 0.95, 0.99*). That is, the likelihood of remaining positive over scaled (*dd*) observation time. Also extract ovrall mean, median and IQR of progression duration time. 

CCDF data is then used to fit a scaled Erlang distribution, which is then re-scaled (just `gamma`) to actual time using overall mean.

```{r}

# Obtain ccdf of PCR positivity
q <- seq(0, 100, l = 21)
df.edata <- data.frame(x = fit.edata$surv,y = fit.edata$cumhaz, z = fit.edata$time)
ccdf_pcr <- data.frame()
for (i in 1:21) {
  x = q[i+1]
  ccdf_pcr[1,i] <- df.edata$y[x]
}
ccdf_pcr <- cbind(0,ccdf_pcr)
ccdf_pcr <- ccdf_pcr[,1:21]
rownames(ccdf_pcr) <- "pcr positive"

# obtain mean, median and IQR of progression durations
progression_durations <- data.frame(param = "m_Com",
                                    Q2.5 = mean(df.edata$z[2],df.edata$z[3]),
                                    Q50 = df.edata$z[50],
                                    Q97.5 = mean(df.edata$z[97],df.edata$z[98]),
                                    mean = mean(df.edata$z))

# pass parameters on to Lilith's functions
source("fit_Erlang.R") ## this is kept offline

# Print outputs
op_res # scaled Erlang parameters (time in dd)
scaled_gamma # re-scaled gamma (true time)


```




